---
title: "Untitled"
author: "Rijo"
date: '2022-10-19'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
getwd()  # Get the current working directory 
setwd('C:/Users/akila/Documents/DS_Masters_R/Foundations for Data Science/FODS_Capstone Datasets/')  # Set the working directory 
RStudio.Version()  # to check for the current version of the RStudio
``` 

# Import the requiredLibraries
```{r}
# Load the required libraries
#install.packages("tidyverse")
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("Hmisc")
#install.packages("tm")
#install.packages("lsa")

#install.packages("treemapify")
#install.packages("patchwork")
library(dplyr)
library(ggplot2)
library(cluster)
library(Hmisc)
library(ggplot2)
library(tidyverse)
library(treemapify)
library(patchwork)
```
### WINTER OLYMPICS

```{r}
# Load the data
olympics_data = read.csv('athlete_events.csv')
#View((olympics_data))

# Filter only by Winter season

winter_subset = filter(olympics_data, Season == 'Winter')
View(winter_subset)   # View the data
str(winter_subset)    # structure of the data 
class(winter_subset)  # class of the data
is.na(winter_subset)  # Check for missing values
colSums(is.na(winter_subset))  # Count the number of missing values per column
which(colSums(is.na(winter_subset))>0) # Identify the position of the columns with atleast one missing value
names(which(colSums(is.na(olympics_data))>0))  # Return the column names with missing values

```
# LIMITATIONS - Winter Subset: 
The original dataset consists of 48564 obs and 15 variables.
The dataset was filtered considering the question in hand. The Winter subset had the following number missing data

"Age"  "Height" "Weight" "Medal" 
 285   8314      9021    42869 

 
# Remove the following columns : 
 i. 'Games' :   The Year and Season column will be used as it is the same data but split into sepearte columns for the ease of analysis
 ii. 'Team' as it has duplicated names for different athletes from the same countries.
 example. Switzerland-1 and Switzerland-2 for Switzerland. The team name has a substitute column 'NOC' which helps in identifying each participating country uniquely.
```{r}
# Remove column 'Team' and 'Games'
selected_winter_variables = select(winter_subset , -(Team),-(Games))
View(selected_winter_variables)
str(selected_winter_variables)

```
# Since 88.27% (42869)of the values are missing in the Medal columns, we filter the data further by Medals
```{r}

# Filtering the Winter subset with Medal
winter_subset_2 = filter(selected_winter_variables, (Season == 'Winter' & Medal %in% c('Gold','Silver','Bronze')))
str(winter_subset_2)

# Count the number of missing values per column
colSums(is.na(winter_subset_2))
# Return the column names with missing values
names(which(colSums(is.na(winter_subset_2))>0))

#View(winter_subset_2)
summary(winter_subset_2) 
```
# IMPUTING MISSING VALUES USING UNIVARIATE IMPUTATION

```{r}

## BEFORE IMPUTATION

# Check if the variables are  normally distributed or not, before proceeding with mean or median
summary(winter_subset_2$Age)
summary(winter_subset_2$Height)
summary(winter_subset_2$Weight)

# GGPLOT RESULTS - Before Imputation

ggplot(data = winter_subset_2, mapping = aes(Age)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Age - Before Imputation')
ggplot(data = winter_subset_2, mapping = aes(Height)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Height - Before Imputation')
ggplot(data = winter_subset_2, mapping = aes(Weight)) + geom_histogram(bins = 40, colour = "black")+ ggtitle('Weight - Before Imputation')


# Apply Univariate Imputation on AGE, HEIGHT and WEIGHT
winter_subset_2$Age = impute(winter_subset_2$Age , fun = mean)
winter_subset_2$Height = impute(winter_subset_2$Height , fun = mean)
winter_subset_2$Weight = impute(winter_subset_2$Weight , fun = mean)

### After Imputation results

summary(winter_subset_2$Age)
summary(winter_subset_2$Height)
summary(winter_subset_2$Weight)

# GGPLOT RESULTS - After Imputation
ggplot(data = winter_subset_2, mapping = aes(Age)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Age - After Imputation')
ggplot(data = winter_subset_2, mapping = aes(Height)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Height - After Imputation')
ggplot(data = winter_subset_2, mapping = aes(Weight)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Weight - After Imputation')


```


# BEFORE IMPUTATION
summary(winter_subset_2$Age)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  13.00   23.00   26.00   26.06   29.00   58.00      11 
summary(winter_subset_2$Height)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  145.0   169.0   176.0   175.5   182.0   201.0     721 
summary(winter_subset_2$Weight)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  36.00   63.00   72.00   72.81   82.00  130.00     837 

# AFTER IMPUTATION
summary(winter_subset_2$Age)

 11 values imputed to 26.06369 

   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  13.00   23.00   26.00   26.06   29.00   58.00 
summary(winter_subset_2$Height)

 721 values imputed to 175.5396 

   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  145.0   170.0   175.5   175.5   181.0   201.0 

summary(winter_subset_2$Weight)

 837 values imputed to 72.80795 

   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  36.00   65.00   72.81   72.81   80.00  130.00 
## The data is now cleaned and has no missing values.

# Variable Type Conversion
```{r}
# 1. Medal: Character -> Oridnal Factor ( Gold, Silver, Bronze)
# 2. Sex : Character -> Nominal Factor


#1. Convert Medal to Ordinal Factor
winter_subset_2$Medal = factor(winter_subset_2$Medal, levels = c('Bronze','Silver','Gold'), ordered = TRUE)

levels(winter_subset_2$Medal)


# 2. Sex : Character to Nominal Factor
winter_subset_2$Sex = factor(winter_subset_2$Sex , levels = c("M","F"))
levels(winter_subset_2$Sex)

str(winter_subset_2)  # View structure of the data after type conversion

```

# Arranging the data from the beginning of the Olympics till date.
```{r}
# sorting the data to check when was the 1st Winter Olympics held
winter_subset_2 = arrange(winter_subset_2, Year)  
View(winter_subset_2)
str(winter_subset_2)
```


### COUNTRY PERFORMANCE ANALYSIS BETWEEN 20TH CENTURY AND 21st CENTURY
```{r}
summarize = dplyr::summarize  # Initialize the summarize function from the dplyr package to be used for summarizing the data.

# Total Medal Count
tapply(winter_subset_2$Medal, winter_subset_2$Medal , length)

# 20th Century 
twentieth_century = winter_subset_2 %>% filter(Year > 1900 & Year < 2000)  # Filter the winter season by the 20th century
#View(twentieth_century)  
str(twentieth_century)


# Distinct medals in the 20th century by NOC(Country) 

gold_20th_century = twentieth_century[twentieth_century$Medal == 'Gold' , c(1:13)] # GOLD
silver_20th_century = twentieth_century[twentieth_century$Medal == 'Silver' , c(1:13)]  # Silver 
bronze_20th_century = twentieth_century[twentieth_century$Medal == 'Bronze' , c(1:13)] # Bronze


## Gold Medal by countries

# 1.Medal Count by Countries - 20th century

gold_medal_by_countries_20th = gold_20th_century %>% group_by(NOC) %>% summarize(Country = n()) %>% arrange(desc(Country))
gold_medal_by_countries_20th

silver_medal_by_countries_20th = silver_20th_century %>% group_by(NOC) %>% summarize(Country = n()) %>% arrange(desc(Country))
silver_medal_by_countries_20th

bronze_medal_by_countries_20th = bronze_20th_century %>% group_by(NOC) %>% summarize(Country = n()) %>% arrange(desc(Country))
bronze_medal_by_countries_20th

## GGPLOT for Gold , Silver and Bronze Medals

# 1. Gold
fig_1 = ggplot(data = head(gold_medal_by_countries_20th,5)) + geom_bar(mapping = aes(x = Country, y = reorder(NOC,Country), fill= NOC), stat = 'identity') + ggtitle("Year 1924-1998 : Gold Medals - Top 5 Countries") + labs( x = 'Gold Medals' ,y = 'Countries')

# 2. Silver
fig_2 = ggplot(data = head(silver_medal_by_countries_20th,5)) + geom_bar(mapping = aes(x = Country, y = reorder(NOC,Country), fill= NOC), stat = 'identity') + ggtitle("Year 1924-1998 : Silver Medals - Top 5 Countries") + labs( x = 'Silver Medals' ,y = 'Countries')

# 3, Bronze
fig_3 = ggplot(data = head(bronze_medal_by_countries_20th,5)) + geom_bar(mapping = aes(x = Country, y = reorder(NOC,Country), fill= NOC), stat = 'identity') + ggtitle("Year 1924-1998 : Bronze Medals - Top 5 Countries") + labs( x = 'Bronze Medals' ,y = 'Countries')

# Combine the plots using patchwork()
fig_1 | fig_2 | fig_3


########################################################################################


# Twenty First Century
twenty_first_century  = winter_subset_2 %>% filter(Year >= 2000)  # Filter the winter season by the 21st century
View(twenty_first_century)
str(twenty_first_century)


# Distinct medals in the 21st century by NOC(Country) 

gold_21st_century = twenty_first_century[twenty_first_century$Medal == 'Gold' , c(1:13)] # GOLD
silver_21st_century = twenty_first_century[twenty_first_century$Medal == 'Silver' , c(1:13)]  # Silver 
bronze_21st_century = twenty_first_century[twenty_first_century$Medal == 'Bronze' , c(1:13)] # Bronze

# 1.Medal Count by Countries - 20th century

#1. Gold 
gold_medal_by_countries_21st = gold_21st_century %>% group_by(NOC) %>% summarize(Country = n()) %>% arrange(desc(Country))
gold_medal_by_countries_21st

#2. Silver
silver_medal_by_countries_21st = silver_21st_century %>% group_by(NOC) %>% summarize(Country = n()) %>% arrange(desc(Country))
silver_medal_by_countries_21st

#3. Bronze
bronze_medal_by_countries_21st = bronze_21st_century %>% group_by(NOC) %>% summarize(Country = n()) %>% arrange(desc(Country))
bronze_medal_by_countries_21st


## GGPLOT's for Gold, Silver , Bronze - 20th Century

#1. GGPLOT - Countries by Gold
fig_4 = ggplot(data = head(gold_medal_by_countries_21st,5)) + geom_bar(mapping = aes(x = Country, y = reorder(NOC,Country), fill= NOC), stat = 'identity') + ggtitle("Year 2002 - 2014 : Gold Medals - Top 5 Countries") + labs( x = 'Gold Medals' ,y = 'Countries')


#2. GGPLOT - Countries by Silver

fig_5 = ggplot(data = head(silver_medal_by_countries_21st,5)) + geom_bar(mapping = aes(x = Country, y = reorder(NOC,Country), fill= NOC), stat = 'identity') + ggtitle("Year 2002 - 2014 : Silver Medals - Top 5 Countries") + labs( x = 'Silver Medals' ,y = 'Countries')

#3. GGPLOT - Countries by Bronze

fig_6 = ggplot(data = head(bronze_medal_by_countries_21st,5)) + geom_bar(mapping = aes(x = Country, y = reorder(NOC,Country), fill= NOC), stat = 'identity') + ggtitle("Year 2002 - 2014 : Bronze Medals - Top 5 Countries") + labs( x = 'Bronze Medals' ,y = 'Countries')

# Combine the plots using patchwork()
fig_4 | fig_5 | fig_6

```


# ATHLETE ANALYSIS - OVERALL

```{r}

str(winter_subset_2)

# Athletes participation(M/F) each year - Grouped by Year

athlete_participation = winter_subset_2 %>% group_by(Year) %>% 
  summarize('Male_Athletes' = sum(Sex == 'M'), 'Female_Athletes' = sum(Sex == 'F'))

View(athlete_participation)


# Scatter Plot : Year- Athlete Participation
#?geom_point

# MALE ATHLETES vs YEAR
SP_1 = ggplot(athlete_participation) +
geom_point(mapping = aes(x = Male_Athletes , y = Year),  size = 5 , pch = 20, color = 'red') + ggtitle("Scatter Plot : Male Athletes vs Year") + xlim(80,340)

# FEMALE ATHLETES vs YEAR
SP_2 = ggplot(athlete_participation) +
geom_point(mapping = aes(x = Female_Athletes , y = Year),   size = 5 , pch = 20, color = 'darkblue') + ggtitle("Scatter Plot : Female Athletes vs Year") + xlim(5,270)

# COMBINING PLOTS using patchwork()
SP_1 | SP_2

```

# ATHLETE - SPORT PARTICIPATION ANALYSIS - 20TH and 21st Century - To determine Sport Popularity

```{r}

# athlete sport participation - 20th century
athlete_sport_participation_20th = twentieth_century %>% group_by(Sport) %>% 
  summarize(Number_of_Athletes = n_distinct(ID)) %>% arrange(desc(Number_of_Athletes))
View(athlete_sport_participation_20th)
 
# athlete sport participation - 21st century
athlete_sport_participation_21st = twenty_first_century %>% group_by(Sport) %>% 
  summarize(Number_of_Athletes = n_distinct(ID)) %>% arrange(desc(Number_of_Athletes))
View(athlete_sport_participation_21st)

## TREE MAP 
# Heat Map - Athlete per Sport - 20th Century
TM_1 = ggplot(athlete_sport_participation_20th, aes(area = Number_of_Athletes, fill = Sport, label = paste(Sport,Number_of_Athletes, sep = '\n'))) + geom_treemap() + geom_treemap_text(colour = "white",place = "centre", size = 20) + theme(legend.position = "none") + labs(title=" 20th Century : Athletes per Sport")  

# Heat Map - Athlete per Sport - 2stth Century
TM_2 = ggplot(athlete_sport_participation_21st, aes(area = Number_of_Athletes, fill = Sport, label = paste(Sport,Number_of_Athletes, sep = '\n'))) + geom_treemap() + geom_treemap_text(colour = "white", place = "centre", size = 20) + theme(legend.position = "none") + labs(title=" 21st Century : Athletes per Sport")  

TM_1 | TM_2

```

# DISTINCT SPORTS 
```{r}
sports_20th = twentieth_century %>% group_by(Sport) %>% 
  summarize(Events = n_distinct(Event)) %>% arrange(desc(Events))
sports_20th

sports_21st = twenty_first_century %>% group_by(Sport) %>% 
  summarize(Events = n_distinct(Event)) %>% arrange(desc(Events))
sports_21st

```
